*Prochlorococcus marinus* responses to light and oxygen

### 2024-02-28 Code/Manuscript_O2.Rmd
Fixed strain name

## *Prochlorococcus marinus* diversity {.unnumbered}
*Prochlorococcus marinus*, a genus of Cyanobacteria, is the smallest known photosynthetic prokaryote, with cell diameters ranging from 0.5 to 0.7 µm [@chisholmProchlorococcusMarinusNov1992]. Despite small cell size, *P. marinus* contribute 13 to 48% of net primary production in oligotrophic oceans, corresponding to about 30% of global oxygen production [@partenskyProchlorococcusMarinePhotosynthetic1999].  *P. marinus* growth is currently limited to between latitudes of 40°N to 40°S in open ocean waters, from surface to 300 m depth, thus spanning 3 orders of magnitudes of light [@partenskyProchlorococcusMarinePhotosynthetic1999; @chisholmProchlorococcusMarinusNov1992 ].

*P. marinus* is functionally differentiated from other cyanobacteria primarily by the absence of phycobilisomes [@chisholmProchlorococcusMarinusNov1992]. Instead, *Prochlorococcus*  uses intra-membrane Prochlorophyte Chlorophyll Binding (Pcb) proteins binding divinylchlorophyll a and b, and α-carotene, as major light harvesting complexes [@goerickePigmentsProchlorococcusMarinus1992]; [@larocheIndependentEvolutionProchlorophyte1996], with diversity in the number of genes encoding Pcb, expression of Pcb under changing light [@garczarekMultiplicationAntennaGenes2000]  [@rocapGenomeDivergenceTwo2003] [@garczarekMultiplicationAntennaGenes2000], and in thylakoid organization [@partenskyProchlorococcusMarinePhotosynthetic1999].  The genes encoding much of the photosynthetic apparatus are highly conserved in *Prochlorococcus* [@kettlerPatternsImplicationsGene2007], though the photosynthetic efficiency may vary among clades [@hessPhotosyntheticApparatusProchlorococcus2001; @moorePhotophysiologyMarineCyanobacterium1999].

*P. marinus* comprises many strains, organized into clades, defined by 16S-23S intergenic transcribed ribosomal sequence signatures [@rocapGenomeDivergenceTwo2003]. The clades inhabit distinct ecological niches [@moorePhysiologyMolecularPhylogeny1998], originally defined by Low-Light (LL) or High-Light (HL).  Of at least 12 known *Prochlorococcus* genetic clades only 5 to date have cultured representatives; HLI, HLII, LLI, LLII/III and LIV [@billerProchlorococcusStructureFunction2015].

@@

*P. marinus* clades are nonetheless found in environments beyond their optimal habitats. HL clades inhabit depths overlapping with LL ecotypes [@westNichePartitioningProchlorococcusPopulationsStratified1999; @zinserInfluenceLightTemperature2007; @delmontLinkingPangenomesMetagenomes2018; @moorePhotophysiologyMarineCyanobacterium1999; @lavinNovelLineagesProchlorococcus2010], while LL clades can occupy regions in OMZ at depths shallower than 40 m, exploiting ambient light levels above what LL clades were thought to tolerate.

@@

Near the equator, photoperiod remains nearly constant at the ocean surface, approximately 12 hours (h) of daylight and 12 h of darkness throughout the year. The effective length of the photoperiod does, however attenuate with depth as dawn and dusk light at depth drops below levels needed for biological processes. As *P. marinus* potentially expands its temperature-permissive niches poleward into temperate regions  [@flombaumPresentFutureGlobal2013; @bartonAnthropogenicClimateChange2016], it will encounter more
pronounced seasonal variations in photoperiod regimes both at surface and at depth, with potentially complex effects upon viable growth niches [@liDiatomGrowthResponses2017; @prezelinDielPeriodicityPhytoplankton1992]. For example, Vaulot *et al*. (1995) @vaulotGrowthProchlorococcusPhotosynthetic1995 showed that *Prochlorococcus* replication of DNA occurs in the afternoon, while cell division occurs at night.  To our knowledge, no study has as yet addressed *P. marinus* growth responses in relation to a  range of photoperiods.

@@

We used the OceansMap Protein Portal (OPP; https://proteinportal.whoi.edu/) [@saitoDevelopmentOceanProtein2021] to analyze the distribution of proteins from clades of *P. marinus* in samples tanken across a range of [O~2~] and depth, which in turn correlates to peak light at the site of sampling. We therefore analyzed the growth and physiological responses of representative strains from three clades of *P. marinus* under a matrix of [O~2~], light levels, spectral waveband ranges, and photoperiods, to approximate eco-physiological conditions representative of current and hypothetical future ocean zones. *P. marinus* MED4, a clade HLI strain, was isolated near the ocean surface (5m depth) of the Mediterranean Sea where [O~2~] is near saturation, light levels are high and spectral bias from full solar irradiance is minimal. *P. marinus* SS120, a clade LLIII strain, was isolated from the Sargasso Sea at a depth of 120 m, while *P. marinus* MIT9313, a clade LLIV strain, was isolated from the North Atlantic Gulf Stream at a depth of 135 m. At these depths, light attenuation and spectral shifts occur, resulting in low blue and green light, while [O~2~] varies from near-surface saturation levels to decreased concentrations, but does not necessarily decrease systematically with depth [@billheimerOxygenSeasonalityUtilization2021].

We then analyzed growth rates in terms of cumulative diel Photosynthetically Usable Radiation (PUR). Photosynthetic organisms absorb light energy within the Photosynthetically Active Radiation (PAR) range, 350 to 700 nm, for photosynthesis [@morelAvailableUsableStored1978]. PUR represents the fraction of PAR that is absorbed by the pigments of a given photosynthetic organism [@morelAvailableUsableStored1978], taking into account the specific spectral wavebands these pigments absorb.  *P. marinus* light-harvesting complexes, [@chisholmProchlorococcusMarinusNov1992;
@goerickePigmentsProchlorococcusMarinus1992], show an absorption maxima
of 442 nm for divinyl chlorophyll a and 478 nm for divinyl chlorophyll b
[@goerickePigmentsProchlorococcusMarinus1992] allowing *P. marinus* to
efficiently harvest blue light in the 400 nm
to 500 nm range [@morelAvailableUsableStored1978] prevailing in deep ocean
habitats, where only blue spectral wavelengths prevail [@holtropVibrationalModesWater2021].  In *P. marinus* small cell diameters, from 0.5 to 0.7 µm [@chisholmProchlorococcusMarinusNov1992], and simple cell structures, minimize the complication of pigment package effect or intracellular self-shading [@morelProchlorococcusSynechococcusComparative1993]. Given the different spectral light regimes typical of the niches of different ecotypes, expressing growth rates in terms of cumulative diel PUR might simplify different photoperiods, spectral bands, and PAR levels into a common parameter, making growth response comparisons across strains and different oxygen levels more accessible.  We also aimed to detect whether growth responses are driven simply by cumulative diel PUR, or whether specific photoperiods, spectral bands or PAR levels have independent, albeit interacting, effects on growth.

>>>>

## *Prochlorococcus marinus* diversity {.unnumbered}
*Prochlorococcus marinus*, a genus of Cyanobacteria, is the smallest known photosynthetic prokaryote, with cell diameters ranging from 0.5 to 0.7 µm [@chisholmProchlorococcusMarinusNov1992]. Despite small cell size, *P. marinus* contribute 13 to 48% of net primary production in oligotrophic oceans, corresponding to about 30% of global oxygen production [@partenskyProchlorococcusMarinePhotosynthetic1999]. *Prochlorococcus marinus* growth is currently limited to between latitudes of 40°N to 40°S in open ocean waters, from surface to 300 m depth, thus spanning 3 orders of magnitudes of light [@partenskyProchlorococcusMarinePhotosynthetic1999; @chisholmProchlorococcusMarinusNov1992 ].

*Prochlorococcus marinus* is functionally differentiated from other cyanobacteria primarily by the absence of phycobilisomes [@chisholmProchlorococcusMarinusNov1992]. Instead, *Prochlorococcus*  uses intra-membrane Prochlorophyte Chlorophyll Binding (Pcb) proteins binding divinylchlorophyll a and b, and α-carotene, as major light harvesting complexes [@goerickePigmentsProchlorococcusMarinus1992]; [@larocheIndependentEvolutionProchlorophyte1996], with diversity in the number of genes encoding Pcb, expression of Pcb under changing light [@garczarekMultiplicationAntennaGenes2000]  [@rocapGenomeDivergenceTwo2003] [@garczarekMultiplicationAntennaGenes2000], and in thylakoid organization [@partenskyProchlorococcusMarinePhotosynthetic1999].  The genes encoding much of the photosynthetic apparatus are highly conserved in *Prochlorococcus* [@kettlerPatternsImplicationsGene2007], though the photosynthetic efficiency may vary among clades [@hessPhotosyntheticApparatusProchlorococcus2001; @moorePhotophysiologyMarineCyanobacterium1999].

*Prochlorococcus marinus* comprises many strains, organized into clades, defined by 16S-23S intergenic transcribed ribosomal sequence signatures [@rocapGenomeDivergenceTwo2003]. The clades inhabit distinct ecological niches [@moorePhysiologyMolecularPhylogeny1998], originally defined by Low-Light (LL) or High-Light (HL).  Of at least 12 known *Prochlorococcus* genetic clades only 5 to date have cultured representatives; HLI, HLII, LLI, LLII/III and LIV [@billerProchlorococcusStructureFunction2015].

@@

*Prochlorococcus marinus* clades are nonetheless found in environments beyond their optimal habitats. HL clades inhabit depths overlapping with LL ecotypes [@westNichePartitioningProchlorococcusPopulationsStratified1999; @zinserInfluenceLightTemperature2007; @delmontLinkingPangenomesMetagenomes2018; @moorePhotophysiologyMarineCyanobacterium1999; @lavinNovelLineagesProchlorococcus2010], while LL clades can occupy regions in OMZ at depths shallower than 40 m, exploiting ambient light levels above what LL clades were thought to tolerate.

@@

Near the equator, photoperiod remains nearly constant at the ocean surface, approximately 12 hours (h) of daylight and 12 h of darkness throughout the year. The effective length of the photoperiod does, however attenuate with depth as dawn and dusk light at depth drops below levels needed for biological processes. As *P. marinus* potentially expands its temperature-permissive niches poleward into temperate regions  [@flombaumPresentFutureGlobal2013; @bartonAnthropogenicClimateChange2016], it will encounter more
pronounced seasonal variations in photoperiod regimes both at surface and at depth, with potentially complex effects upon viable growth niches [@liDiatomGrowthResponses2017; @prezelinDielPeriodicityPhytoplankton1992]. For example, Vaulot *et al*. @vaulotGrowthProchlorococcusPhotosynthetic1995 showed that *Prochlorococcus* replication of DNA occurs in the afternoon, while cell division occurs at night.  To our knowledge, no study has as yet addressed *P. marinus* growth responses in relation to a  range of photoperiods.

@@

We used the OceansMap Protein Portal (OPP; https://proteinportal.whoi.edu/) [@saitoDevelopmentOceanProtein2021] to analyze the distribution of proteins from clades of *P. marinus* in samples tanken across a range of [O~2~] and depth, which in turn correlates to peak light at the site of sampling. We therefore analyzed the growth and physiological responses of representative strains from three clades of *P. marinus* under a matrix of [O~2~], light levels, spectral waveband ranges, and photoperiods, to approximate eco-physiological conditions representative of current and hypothetical future ocean zones. *Prochlorococcus marinus* MED4, a clade HLI strain, was isolated near the ocean surface (5m depth) of the Mediterranean Sea where [O~2~] is near saturation, light levels are high and spectral bias from full solar irradiance is minimal. *Prochlorococcus marinus* SS120, a clade LLIII strain, was isolated from the Sargasso Sea at a depth of 120 m, while *P. marinus* MIT9313, a clade LLIV strain, was isolated from the North Atlantic Gulf Stream at a depth of 135 m. At these depths, light attenuation and spectral shifts occur, resulting in low blue and green light, while [O~2~] varies from near-surface saturation levels to decreased concentrations, but does not necessarily decrease systematically with depth [@billheimerOxygenSeasonalityUtilization2021].

We then analyzed growth rates in terms of cumulative diel Photosynthetically Usable Radiation (PUR). Photosynthetic organisms absorb light energy within the Photosynthetically Active Radiation (PAR) range, 350 to 700 nm, for photosynthesis [@morelAvailableUsableStored1978]. PUR represents the fraction of PAR that is absorbed by the pigments of a given photosynthetic organism [@morelAvailableUsableStored1978], taking into account the specific spectral wavebands these pigments absorb. *Prochlorococcus marinus* light-harvesting complexes, [@chisholmProchlorococcusMarinusNov1992;
@goerickePigmentsProchlorococcusMarinus1992], show an absorption maxima
of 442 nm for divinyl chlorophyll a and 478 nm for divinyl chlorophyll b
[@goerickePigmentsProchlorococcusMarinus1992] allowing *P. marinus* to
efficiently harvest blue light in the 400 nm
to 500 nm range [@morelAvailableUsableStored1978] prevailing in deep ocean
habitats, where only blue spectral wavelengths prevail [@holtropVibrationalModesWater2021].  In *P. marinus* small cell diameters, from 0.5 to 0.7 µm [@chisholmProchlorococcusMarinusNov1992], and simple cell structures, minimize the complication of pigment package effect or intracellular self-shading [@morelProchlorococcusSynechococcusComparative1993]. Given the different spectral light regimes typical of the niches of different ecotypes, expressing growth rates in terms of cumulative diel PUR might simplify different photoperiods, spectral bands, and PAR levels into a common parameter, making growth response comparisons across strains and different oxygen levels more accessible.  We also aimed to detect whether growth responses are driven simply by cumulative diel PUR, or whether specific photoperiods, spectral bands or PAR levels have independent, albeit interacting, effects on growth.
